
\begin{thebibliography}{213}

\addcontentsline{toc}{chapter}{Литература}

\English

\bibitem{1}
Harold Abelson, Gerald Jay Sussman, and Julie Sussman. \emph{Structure and Interpretation of Computer Programs}. The MIT Press, Cambridge, Mass, 1985.

\bibitem{2}
Harold Abelson, Gerald Jay Sussman, and Julie Sussman. \emph{Structure and Interpretation of Computer Programs, Second Edition.} The MIT Press, Cambridge, Mass, 1996.

\bibitem{3}
Ili\`es Alouini and Peter Van Roy. Le protocole r\'eparti du langage Distributed Oz (The distributed protocol of the Distributed Oz language). In \emph{Colloque Francophone d’Ing\'enierie de Protocoles (CFIP 99)}, pages 283–298, April 1999.

\bibitem{4}
Edward G. Amoroso. \emph{Fundamentals of Computer Security Technology}. Prentice Hall, 1994.

\bibitem{5}
Ross J. Anderson. \emph{Security Engineering: A Guide to Building Dependable Distributed Systems.} John Wiley \& Sons, 2001.

\bibitem{6}
Gregory R. Andrews. \emph{Concurrent Programming: Principles and Practice}. Addison-Wesley, 1991.

\bibitem{7}
Joe Armstrong. Higher-order processes in Erlang, January 1997. Unpublished talk.

\bibitem{8}
Joe Armstrong. Concurrency oriented programming in Erlang, November 2002. Invited talk, Lightweight Languages Workshop 2002.

\bibitem{9}
Joe Armstrong, Mike Williams, Claes Wikstr\"om, and Robert Virding. \emph{Concurrent Programming in Erlang.} Prentice-Hall, Englewood Cliffs, N.J., 1996.

\bibitem{10}
Ken Arnold and James Gosling. \emph{The Java Programming Language, Second Edition.} Addison-Wesley, 1998.

\bibitem{11}
Arvind and R. E. Thomas. I-Structures: An efficient data type for functional languages. Technical Report 210, MIT, Laboratory for Computer Science, 1980.

\bibitem{12}
John Backus. Can programming be liberated from the von Neumann style? A functional style and its algebra of programs. \emph{Communications of the ACM}, 21(8):613–641, August 1978.

\bibitem{13}
John Backus. The history of FORTRAN I, II and III. \emph{ACM SIGPLAN Notices}, 13(8), August 1978.

\bibitem{14}
Henri E. Bal, Jennifer G. Steiner, and Andrew S. Tanenbaum. Programming languages for distributed computing systems. \emph{ACM Computing Surveys}, 21(3):261–322, September 1989.

\bibitem{15}
Holger B\"ar, Markus Bauer, Oliver Ciupke, Serge Demeyer, St\'ephane Ducasse, Michele Lanza, Radu Marinescu, Robb Nebbe, Oscar Nierstrasz, Michael Przybilski, Tamar Richner, Matthias Rieger, Claudio Riva, Anne-Marie Sassen, Benedikt Schulz, Patrick Steyaert, Sander Tichelaar, and Joachim Weisbrod. \emph{The FAMOOS Object-Oriented Reengineering Handbook.} October 1999. Result of ESPRIT project FAMOOS.

\bibitem{16}
Philip A. Bernstein, Vassos Hadzilacos, and Nathan Goodman. \emph{Concurrency Control and Recovery In Database Systems.} Addison-Wesley, 1987.

\bibitem{17}
Richard Bird. \emph{Introduction to Functional Programming using Haskell, Second Edition.} Prentice Hall, 1998.

\bibitem{18}
Andrew D. Birrell and Bruce Jay Nelson. Implementing remote procedure calls. \emph{ACM Transactions on Computer Systems}, 2(1):39–59, February 1984.

\bibitem{19}
Darius Blasband. Language engineering: from a hobby, to a research activity, to a trade, March 2002. Unpublished talk.

\bibitem{20}
Per Brand, Peter Van Roy, Rapha\"el Collet, and Erik Klintskog. Path redundancy in a mobile-state protocol as a primitive for language-based fault tolerance. Technical Report RR2000-01, D\'epartement d’Ing\'enierie Informatique, Universit\'e catholique de Louvain, 2000. Available at \verb"http://www.info.ucl.ac.be".

\bibitem{21}
Ivan Bratko. \emph{PROLOG Programming for Articial Intelligence, Third Edition}. Addison-Wesley, 2000.

\bibitem{22}
Per Brinch Hansen. Structured multiprogramming. \emph{Communications of the ACM}, 15(7):574–578, July 1972.

\bibitem{23}
Per Brinch Hansen. \emph{Operating System Principles}. Prentice Hall, 1973.

\bibitem{24}
Per Brinch Hansen. Java’s insecure parallelism. \emph{ACM SIGPLAN Notices}, 34(4):38–45, April 1999.

\bibitem{25}
Frederick P. Brooks, Jr. \emph{The Mythical Man-Month: Essays on Software Engineering}. Addison-Wesley, 1975.

\end{thebibliography}
